 

### 2021-06-28 vertical-thermal-review.Rmd
working on in-situ observations section
@NidhiVinod , I'm confused by this section and hope I'm not making a mess of it. I'll open an issue on this.

### In-situ observations
**Following the principles outlined above, vertical microclimate profile and leaf traits (see following section) together shape in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
Canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.
Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. Upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig. `r fig_NEON_vertical`). At a specie-specific level, leaf boundary resistance is lower for conifers than broad-leaf species, resulting in higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018]. Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].

In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
Under other conditions $T_{leaf}$ tracks $T_{air}$, due to high evapotranspiration and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

>>>>

### In-situ observations
**Following the principles outlined above, vertical microclimate profile and leaf traits (see following section) together explain in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is commonly greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
Under other conditions $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s#* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

The extent of this buffering varies across forest types.
*Upper canopies of temperate-deciduous species can have greater $T_{leaf} - T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig. `r fig_NEON_vertical`).* 
Because conifer leaves have lower $g_a$ than broad leaves, conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018]. 
Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
